Omics-based Single Step Trait Prediction

### 2017-03-29 paper/manuscript.tex
Expand discussion, add results on manual core sampling

%% THE USEPACKAGES NECESSARY FOR THIS EXAMPLE
%% NOTE THAT genetics_manu_style MUST BE CALLED AFTER mychicago
\usepackage{graphicx}
\usepackage{endfloat}
\usepackage{amsfonts}
\usepackage{mychicago}
\usepackage{subfigure}
\usepackage{genetics_manu_style}
\usepackage{authblk} % improved citations
\usepackage{setspace} %double spacing
\usepackage{lineno} % line numbers
\usepackage{booktabs} % book-quality tables
\usepackage{multicol} % list corresponding authors in two separate columns

@@

% Introduction
\section{Introduction} 
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by accelerating the selection of promising candidates 
without phenotypes \cite{DeLosCampos2013,Garcia-Ruiz2016} and recently emerged
as an important tool in human diagnostics \cite{DeLosCampos2010,Vazquez2016}.
Its use is largely motivated by the high costs and oftentimes long duration of 
phenotyping, which can be reduced when covering all genotypes with a predictor 
while limiting phenotyping to a few genotypes with high precision.
The performance of all other genotypes is then forecast using model parameters
obtained from thorough (repeated) $k$-fold cross-validation routines applied to 
the subset of individuals with full information.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50-100\% for yield traits and by 300-400\% for traits with low 
heritability \cite{Garcia-Ruiz2016}.
In plant breeding, the relative excellence of a breeding strategy largely 
depends on the prediction accuracy, which may vary widely among traits and 
crops \cite{Longin2015}.

% Search for better predictors
Compared to pedigree information, which represent the expected relationship
among individuals, genomic information capture Mendelian sampling and thereby 
provide an improved proxy of the realized relationship among individuals.
Nevertheless, genomic data do not exhaustively capture physiological epistasis 
\cite{Jiang2015,Guo2016,Vazquez2016}, describing the interactions 
within and between different biological strata \cite{Sackton2016}.
Such interactions were found to be pervasive throughout the genomes of yeast 
\cite{Brem2005} and humans \cite{Brown2014} and have motivated studies on the 
utility of downstream 'omics' predictors for integrating such interactions.
So far, encouraging results have been found for humans \cite{Vazquez2016}, maize 
inbred lines \cite{Guo2016} and maize hybrids \cite{Westhues2017}.

% Incomplete predictors.
Breeding programs --- particularly in animal breeding --- are characterized by
the availability of one predictor type, which is available for all recorded
genotypes while other predictors are incomplete in that they cover only a 
subset of all genotypes.
Usually, the complete predictor is pedigree information while genomic data are
the most prevalent incomplete predictor.
To utilize all available phenotypic information for model training, methods were 
needed that would allow for a combined use of genomic and pedigree information 
\cite{Hayes2009a,VanRaden2009}.

% Two-step prediction.
In the early years of genomic prediction, a two-step procedure was the most
intuitive choice for establishing a new selection scheme.
The first step constists of conventionally estimating breeding values based on
pedigree relationships and the phenotypes of relatives.
Individuals with accurate breeding values out of the first step (\textit{i.e.}
high number of offspring), enter the second step in which conventional breeding
values are regressed onto SNP genotypes.
The resulting model coefficients can then be used to predict genomic breeding 
values for solely genotyped and potentially very young individuals.
However, as the breeding values from step one enter the genomic prediction
model as phenotypes, one has to be careful about the residual error
distribution of such a model.
For instance, if the range of the prediction errors will not be identically
distributed anymore, \textit{i.e.} they should be weighted differently or
estimated individually \cite{Aguilar2010}.
One popular approach to account for this problem is deregressing the breeding
values and weighting the residuals according to the prediction error variances
\cite{Garrick2009}.

@@

% -   Mention scarcity of data on single step prediction in plant populations.
Single step prediction has already become a routine procedure in animal
breeding but, to the best of our knowledge, has only been considered in plants
in a study on wheat lines \cite{Ashraf2016}. 
Our objectives were threefold: (1) Examine whether the single step prediction
framework can be transfered to plant hybrids, (2) explore the utility of single 
step predition with a quantitative predictor and (3) evaluate the impact of the
genetic space covered by the incomplete predictor on predictive ability.

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
The maize hybrid data set comprises 1,521 hybrids produced in 16 factorial
mating designs between 142 Dent and 103 Flint parent lines \cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits.
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper \cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50 \cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software \cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for parental Dent and
6,212 for parental Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines through two-color
hybridizations using a custom 2K microarray (GPL22267) \cite{Westhues2017}.
After applying established normalization procedures 
\cite{Smyth2003,Ritchie2007}, gene expression BLUEs were computed for 1,323 
transcripts in reference to established protocols 
\cite{Smyth2003,Ritchie2007,Frisch2010} using the R-package \textit{limma}
\cite{Ritchie2015a}.

@@

\Genetics2level{Core sampling}
In order to evaluate the influence of the genetic constitution of the set of
genotypes that has only data on one out of two predictors, we generated core 
samples using the R-package \textit{corehunter} \cite{Thachuk2009}.
These core samples were based on the 149 maize inbred lines from the diversity
panel which had gene expression as well as genomic information.
The size of the core set was varied in increments of ten percentage points and 
ranged from 10\% to 90\% of all 149 inbred lines.
For assembling the various core sets, we first computed the Modifed Rogers (MR)
distance between each pair of genotypes and then selected a predefined subset
of genotypes that display maximum average distance between each other.

@@

\subsection{Universal breeding value model}
The universal model for breeding values was:

@@

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, $\mathbf{Z}_{lo}$ is the 
corresponding design matrix associating the random genotype effects 
(\textit{i.e.}, breeding values) of the lines in the $l$-th inbred line group 
and the $o$-th predictor type with $\mathbf{y}$.
Note that, in the case of hybrid data, $\mathbf{Z}_{lo}$ associates the random
general combining ability effects in the parental inbred lines (either Dent or
Flint, as expressed by $l$) with the vector of observed hybrid performance
$\mathbf{y}$.
The random effects ($\mathbf{g}_{lo}$) have expectation zero and variance equal 
to $\mathbf{K}_{lo} \sigma^{2}_{{g}^{lo}}$ and $\mathbf{I} \sigma^2_{\epsilon}$ 
for the residual error.
All prediction models were implemented using the R-package \textit{BGLR}
\cite{Perez2014}.

\subsection{Single step prediction}
The breeding values in a prediction model are in general given by:
\begin{equation} \label{eq:mrnaebv}
	\mathbf{\tilde{g}} = \mathbf{W}\boldsymbol{\hat{\alpha}},
\end{equation}

@@

\begin{equation} \label{eq:single-step-model}
\mathbf{y} = \mathbf{Xb} + \mathbf{W} \boldsymbol{\alpha} + \mathbf{U}^{Complete} \boldsymbol{\epsilon}^{Complete} + \mathbf{e},
\end{equation}

@@

\begin{equation} \label{eq:single-step-submatrices}
\mathbf{X} = 
\begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix},
 \mathbf{W} = 
\begin{bmatrix}
  Z_1\hat{\mathbf{W}_1} \\
  \mathbf{W}_2 
 \end{bmatrix},
 \mathbf{U}^{Complete} = 
\begin{bmatrix}
  Z_1 \\
  0 
 \end{bmatrix}
\end{equation}

@@

\begin{align} \label{eq:entangled-augmented-single-step-model}
\begin{bmatrix}
  y_1 \\
  y_2 
 \end{bmatrix}
& =
 \begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_1 & 0 \\
  0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  g_1 \\
  g_2 
 \end{bmatrix}
  + \mathbf{e} \\
    & = 
 \begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_1 & 0 \\
  0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{K}_{12}\mathbf{K}_{22}^{-1}\mathbf{W}_2\boldsymbol{\alpha} + \boldsymbol{\epsilon}^{Complete}  \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
    & = 
 \begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_1 & 0 \\
  0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  \hat{\mathbf{W}_1}\boldsymbol{\alpha} + \boldsymbol{\epsilon}^{Complete} \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
\end{align}

The breeding values from that model are:

\begin{equation} \label{eq:breeding-values}
\tilde{\mathbf{g}} = 
 \begin{bmatrix}
  \hat{\mathbf{W}_1} \\
  \mathbf{W}_2 
 \end{bmatrix}
 \hat{\boldsymbol{\alpha}}
 + 
 \begin{bmatrix}
  \mathbf{Z}_1 \\
  0
 \end{bmatrix}
 \hat{\boldsymbol{\epsilon}}^{Complete} 
\end{equation}

@@

\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\sum_{l=1}^{L} \mathbf{Z}_{l}\mathbf{W}_{l} \boldsymbol{\alpha}_{l} + 
\sum_{l=1}^{L} \mathbf{Z}_{l}\mathbf{U}_{l}^{Complete} \boldsymbol{\epsilon}_{l}^{Complete} +
\mathbf{e}.
\end{align}

@@

\begin{equation} \label{eq:predicted-performance}
\hat{\mathbf{y}} = \sum_{l=1}^{L} \mathbf{Z}_{l}\tilde{\mathbf{g}}_{l} 
\end{equation}

@@

where $n$ denotes the number of genotypes (*i.e.*, inbred lines or hybrids).

\section*{Results}
\Genetics2level{Inbred lines}
At the level of the 149 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for four out of six traits
(Table \ref{table:InbredResults}).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.04 as observed for the traits '100grainweight', 'Eardiameter' and 
'Silkingtime'.
Except for 'Kernelwidth', the predictive ability for predictions based on
G and the full set of inbred lines ($n = 211$) was always higher than for the 
reduced set of inbred lines ($n = 149$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 211 lines performed at least as well as G alone when T was 
superior over G for the reduced set of 149 lines.
In the case of '100grainweight' and 'Eardiameter', the use of GT yielded a 
slight improvement in predictive ability over G.
Where T was a worse predictor for the reduced set of 149 lines than G,
predictive abilities of GT for the full set of 211 lines were also worse than 
those based on G alone.

% Script: ./analysis/maizego_core_fraction_plot_predictive_abilities.R
\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/inbred_line_combi_plot.pdf}
  \caption{
  Predictive abilities and corresponding coefficients of variation for 
  different core sets from the tropical/subtropical material of the maize 
  diversity panel.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  The decimal after the predictor name indicates the size of the genetic 
  material that was covered by inbred lines in the core set.
  Influence of core set composition on predictive abilities.
  Predictive abilities are based on single step predition with genomic data as the 
  complete predictor and transcriptomic data as the incomplete predictor.
  Results are shown for six agronomic traits and nine core sets of the maize 
  diversity panel.
  } 
  \label{fig:InbredResults}
\end{figure}

\Genetics2level{Hybrids}
The maize hybrid data set contains information on six agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the 'Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as 'Core'.
For dry matter yield (DMY), T was the best predictor in the core set of 685 
hybrids with a predictive ability that was 0.08 points higher than for genomic 
data, as the second best predictor for this trait 
(Table \ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for starch ('STA').
For the other four traits, G had the best predictive abilities.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of 'FAT'.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement of G alone for the trait 'DMY'.
All traits, where T was not superior to G in the core set, did not benefit from 
the single-step approach when compared to genomic information alone.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the core set for four traits.

% latex table generated in R 3.3.3 by xtable 1.8-2 package
% Fri Mar 24 12:24:30 2017
% Script: ./analysis/uhoh_plot_predictive_abilities.R
\begin{figure}[H]
\centering
  \includegraphics{./tables_figures/hybrid_predictive_ability.pdf}
  \caption{
    Predictive abilities and corresponding coefficients of variation for the 
    set of maize hybrids.
    As predictors, pedigree (P), genomic (G), transcriptomic (T) data and 
    combinations thereof we used.
  } 
  \label{fig:HybridResults}
\end{figure}

\Genetics2level{Coverage of the genetic space}
Single-step approaches are based on the use of one complete predictor that
covers all genotypes and one incomplete predictor that covers only a subset of
all available genotypes.
Here, we explore the impact of which data are covered by the incomplete 
predictor on the predictive abilities obtained via single-step approaches by
focusing exclusively on the 149 inbred lines from the maize diversity panel
that are covered by both, genomic and transcriptomic information.
First, we assmbled nine different core sets, which we defined as subsets
of genotypes that maximize the average genetic distance between genotypes of
the full set.
The nine core sets differ in that each covers a different number of genotypes,
raning from only 10\% to to 90\% of all genotypes covered by the incomplete
predictor (Fig. \ref{fig:CoreSetPCA}).

@@

Predictive abilities obtained with the different core sets varied widely within
traits (Fig. \ref{fig:InbredResults}C).
For the traits 'cobweight' and 'Plantheight', which were better predicted via G
than via T in the reduced set of lines, predictive abilities were higher when
only a small number of genotypes was included in the core set
For most other traits, an initial increase in predictive abilities when moving
from a core set size of 0.1 to 0.5/0.6 was observed, followed by a drop around
0.6/0.7 and a subsequent hike up to a core set size of 0.9.

@@

Of the 149 genotypes, 17 belonged to more than 50\% to the first ancestral
population (A1), 109 belonged to more than 50\% to the second ancestral
population (A2) and 19 belonged to more than 50\% to the third ancestral
population while four genotypes could not be unequivocally assigned to either
ancestral population (mixed).
To adjust the size of the set of genotypes largely belonging to A2 to the other
two sets, we selected 19 genotypes from A2 using previously described core
sampling.
For single-step prediction, each of the three genotype sets were then declared
as having genomic and transcriptomic information whereas their complements were
covered by transcriptomic data only.
Genotypes from A1 and A3, which were at the far ends in the PCA plot (Fig.
\ref{fig:PopStructure}B), were estimated with less precision than genotypes
from the central cluster of genotypes belonging largely to the A2 population
(Fig. \ref{fig:PopStructure}A, Table \ref{fig:InbredTable}).

>>>>

%% THE USEPACKAGES NECESSARY FOR THIS EXAMPLE
%% NOTE THAT genetics_manu_style MUST BE CALLED AFTER mychicago
\usepackage{graphicx}
\usepackage{endfloat}
\usepackage{amsfonts}
\usepackage{mychicago}
\usepackage{subfigure}
\usepackage{genetics_manu_style}
\usepackage{authblk} % improved citations
\usepackage{setspace} %double spacing
\usepackage{lineno} % line numbers
\usepackage{booktabs} % book-quality tables
\usepackage{multicol} % list corresponding authors in two separate columns
\usepackage{hyperref} % web links

@@

% Introduction
\section{Introduction} 
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by accelerating the selection of promising candidates 
without phenotypes \cite{DeLosCampos2013,Garcia-Ruiz2016} and recently emerged
as an important tool in human diagnostics \cite{DeLosCampos2010,Vazquez2016}.
Its use is largely motivated by the high costs and oftentimes long duration of 
phenotyping, which can be reduced when covering all genotypes with a predictor 
while limiting phenotyping to a few genotypes with high precision
\cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
obtained from thorough (repeated) $k$-fold cross-validation routines applied to 
the subset of individuals with full information \cite{Technow2014}.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50-100\% for yield traits and by 300-400\% for traits with low 
heritability \cite{Garcia-Ruiz2016}.
In plant breeding, the relative excellence of a breeding strategy largely 
depends on the prediction accuracy, which may vary widely among traits and 
crops \cite{Longin2015}.

% Incomplete predictors.
Breeding programs --- particularly in animal breeding --- are characterized by
the availability of one predictor type, which is available for all recorded
genotypes while other predictors are incomplete in that they cover only a 
subset of all genotypes \cite{Fragomeni2015}.
Usually, the complete predictor is pedigree information while genomic data are
the most prevalent incomplete predictor.
To utilize all available phenotypic information for model training, methods were 
needed that would allow for a combined use of genomic and pedigree information 
\cite{Hayes2009a,VanRaden2009}.

% Two-step prediction.
In the early years of genomic prediction, a two-step procedure was the most
intuitive choice for establishing a new selection scheme \cite{VanRaden2009}.
The first step consists of conventionally estimating breeding values based on
pedigree relationships and the phenotypes of relatives.
Individuals with accurate breeding values out of the first step (\textit{i.e.}
high number of offspring), enter the second step in which conventional breeding
values are regressed onto SNP genotypes.
The resulting model coefficients can then be used to predict genomic breeding 
values for solely genotyped and potentially very young individuals.
However, as the breeding values from step one enter the genomic prediction
model as phenotypes, one has to be careful about the residual error
distribution of such a model.
For instance, if the range of the prediction errors will not be identically
distributed anymore, \textit{i.e.} they should be weighted differently or
estimated individually \cite{Aguilar2010}.
One popular approach to account for this problem is deregressing the breeding
values and weighting the residuals according to the prediction error variances
\cite{Garrick2009}.

@@

% -   Mention scarcity of data on single step prediction in plant populations.
In animal breeding, single step prediction has already become a routine 
procedure \cite{Legarra2014} but, to the best of our knowledge, has only been 
considered in plants in a study on wheat lines \cite{Ashraf2016}. 
Our objectives were threefold: (1) Examine whether the single step prediction
framework can be transfered to plant hybrids, (2) explore the utility of single 
step predition with a quantitative predictor and (3) evaluate the impact of the
genetic space covered by the incomplete predictor on predictive ability.

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
The maize hybrid data set comprises 1,521 hybrids produced in 16 factorial
mating designs between 142 Dent and 103 Flint parent lines \cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits.
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper \cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50 \cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software \cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for parental Dent and
6,212 for parental Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny, through two-color hybridizations using a custom 2K microarray 
(GPL22267) \cite{Westhues2017}.
After applying established normalization procedures 
\cite{Smyth2003,Ritchie2007}, gene expression BLUEs were computed for 1,323 
transcripts in reference to established protocols 
\cite{Smyth2003,Ritchie2007,Frisch2010} using the R-package \textit{limma}
\cite{Ritchie2015a}.

@@

\Genetics2level{Core sampling}
In order to evaluate the influence of the genetic constitution of the set of
genotypes that has only data on one out of two predictors, we generated core
samples based on two different approaches.
All core sets were assembled from the 149 maize inbred lines from the diversity
panel which had gene expression as well as genomic information.
We defined core samples as a subset of genotypes that is covered by two
predictors whereas all other genotypes (\textit{i.e.}, the complement) are only
covered by a single predictor.
The first application involved assembling core sets of varying sizes by 
maximizing the average genetic distance among core set members using the 
Modified Rogers (MR) distance as the criterion \cite{Thachuk2009}.
The size of the core set was varied in increments of ten percentage points and 
ranged from 10\% to 90\% of all 149 inbred lines.
For the second application, we first ran an admixture analysis in combination
with a PCA and kept $K=3$ as the number of putative ancestral population, which 
seemed to describe the population structure best.
Genotypes were then assigned to one of the three ancestral populations based on
whether any of the three relationship coefficients (A1-A3) was equal to or
greater than 0.5.
For example, a genotype with an admixture coefficient for A2 $\geq 0.5$ was
assigned to the second ancestral population.
Thereby, we obtained $n_{A1} = 17$, $n_{A2} = 109$, $n_{A3} = 19$ and four 
genotypes that could not be unambiguously assigned to any putative ancestral 
population based on this criterion.
To ensure commensurability of the three subpopulations within the set of 149
genotypes, the subset of 109 genotypes (\textit{i.e.}, members of A2) was
reduced to $n_{A2} = 19$ genotypes, again using the method of maximizing the
average genetic distance among selected genotypes described above.

@@

\subsection{Breeding value model}
The general model for breeding values was:

@@

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, $\mathbf{Z}_{lo}$ is the 
corresponding design matrix associating the random genotype effects
$\mathbf{g}_{lo}$ (\textit{i.e.}, breeding values) of the lines in the $l$-th 
inbred line group and the $o$-th predictor type with $\mathbf{y}$.
Note that, in the case of hybrid data, $\mathbf{g}_{lo}$ pertains to the
general combining ability effects in the parental inbred lines (either Dent or
Flint, as expressed by $l$) which are associated with the vector of
observed hybrid performance $\mathbf{y}$ through $\mathbf{Z}_{lo}$.
The random effects ($\mathbf{g}_{lo}$) have expectation zero and variance equal 
to $\mathbf{K}_{lo} \sigma^{2}_{{g}^{lo}}$ and $\mathbf{I} \sigma^2_{\epsilon}$ 
for the residual error.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

\subsection{Single step prediction}
The breeding values in a prediction model are in general given by:
\begin{equation} \label{eq:mrnaebv}
	\mathbf{\hat{g}} = \mathbf{W}\boldsymbol{\hat{\alpha}},
\end{equation}

@@

\begin{equation} \label{eq:single-step-model}
\mathbf{y} = \mathbf{Xb} + \mathbf{W} \boldsymbol{\alpha} + \mathbf{U} \boldsymbol{\epsilon} + \mathbf{e},
\end{equation}

@@

\begin{equation} \label{eq:single-step-submatrices}
\mathbf{X} = 
\begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix},
 \mathbf{W} = 
\begin{bmatrix}
  Z_1\hat{\mathbf{W}_1} \\
  \mathbf{W}_2 
 \end{bmatrix},
 \mathbf{U} = 
\begin{bmatrix}
  Z_1 \\
  0 
 \end{bmatrix}
\end{equation}

@@

\begin{align} \label{eq:entangled-augmented-single-step-model}
\begin{bmatrix}
  y_1 \\
  y_2 
 \end{bmatrix}
& =
 \begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_1 & 0 \\
  0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  g_1 \\
  g_2 
 \end{bmatrix}
  + \mathbf{e} \\
    & = 
 \begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_1 & 0 \\
  0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{K}_{12}\mathbf{K}_{22}^{-1}\mathbf{W}_2\boldsymbol{\alpha} + \boldsymbol{\epsilon}  \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
    & = 
 \begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_1 & 0 \\
  0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  \hat{\mathbf{W}_1}\boldsymbol{\alpha} + \boldsymbol{\epsilon} \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
\end{align}

@@

\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\sum_{l=1}^{L} \mathbf{Z}_{l}\mathbf{W}_{l} \boldsymbol{\alpha}_{l} + 
\sum_{l=1}^{L} \mathbf{Z}_{l}\mathbf{U}_{l} \boldsymbol{\epsilon}_{l} +
\mathbf{e}.
\end{align}

Breeding values can be estimated as:

\begin{equation} \label{eq:breeding-values}
\hat{\mathbf{g}} = 
 \begin{bmatrix}
  \hat{\mathbf{W}_1} \\
  \mathbf{W}_2 
 \end{bmatrix}
 \hat{\boldsymbol{\alpha}}
 + 
 \begin{bmatrix}
  \mathbf{Z}_1 \\
  0
 \end{bmatrix}
 \hat{\boldsymbol{\epsilon}} 
\end{equation}

@@

\begin{equation} \label{eq:predicted-performance}
\hat{\mathbf{y}} = \sum_{l=1}^{L} \mathbf{Z}_{l}\hat{\mathbf{g}}_{l} 
\end{equation}

@@

where $n$ denotes the number of genotypes (\textit{i.e.}, inbred lines or 
hybrids).

\section*{Results}
\Genetics2level{Inbred lines}
At the level of the 149 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for four out of six traits
(Fig. \ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.04 as observed for the traits DS, ED, GW and KW.
Except for KW, the predictive ability for predictions based on G were always
higher for the full set of inbred lines ($n = 211$) than for the reduced set of 
inbred lines ($n = 149$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 211 lines performed at least as well as G alone when T was 
superior over G for the reduced set of 149 lines.
In the case of GW and ED, the use of GT yielded a slight improvement in predictive 
ability over G.
Where T was a worse predictor for the reduced set of 149 lines than G,
predictive abilities of GT for the full set of 211 lines were also worse than 
those based on G alone.

% Script: ./analysis/maizego_core_fraction_plot_predictive_abilities.R
\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/inbred_line_combi_plot.pdf}
  \caption{
  Predictive abilities and average coefficient of variation for
  tropical/subtropical inbred lines from the maize diversity panel and six 
  agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  (\textbf{A}) The 'Core' panel includes a subset of 149 lines covered by
  genomic and transcriptomic information whereas the 'Full' panel comprises all
  211 genotypes. In the 'Full' set of genotypes, transcriptomic records were
  imputed for 62 lines.
  (\textbf{B}) Influence of structural missingness in the coverage of genotypes 
  with transcriptomic information on single step-based predictive abilities 
  using a 'Core' set of 149 inbred lines.
  Transcriptomic records were deleted for 17 to 19 genotypes that were
  primarily associated with a particular subgroup in the material, as defined
  by an admixture analysis and a PCA (Fig. \ref{fig:PopStructure}).
  (\textbf{C}) Influence of core set sampling on predictive abilities using
  nine different sub set sizes where 'Core Fraction' indicates the fraction
  of 149 inbred lines that is covered by both, genomic and transcriptomic
  data.
  } 
  \label{fig:InbredResults}
\end{figure}

\Genetics2level{Hybrids}
The maize hybrid data set contains information on six agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the 'Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as 'Core'.
For dry matter yield (DMY), T was the best predictor in the core set of 685 
hybrids with a predictive ability that was 0.08 points higher than for genomic 
data, as the second best predictor for this trait 
(Fig. \ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for starch ('STA') and 
dry matter content ('DMC').
For the other three traits, inside the core set G was the superior predictor.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of 'FAT'.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement of G alone for the trait 'DMY'.
All traits, where T was not superior to G in the core set, did not benefit from 
the single-step approach when compared to genomic information alone.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the core set for four traits.

% Script: ./analysis/uhoh_plot_predictive_abilities.R
\begin{figure}[H]
\centering
  \includegraphics{./tables_figures/hybrid_predictive_ability.pdf}
  \caption{
    Predictive abilities and average coefficients of variation for the set of 
    maize hybrids and six agronomic traits.
    As predictors, pedigree (P), genomic (G), transcriptomic (T) data and 
    combinations thereof we used.
    The 'Full' set of genotypes comprises 1,521 hybrid progeny based on 245
    parental inbred lines, whereas the 'Core' set comprises 685 hybrid progeny
    based on 103 parental inbred lines.
  } 
  \label{fig:HybridResults}
\end{figure}

\Genetics2level{Coverage of the genetic space}
Single-step approaches are based on the use of one complete predictor that
covers all genotypes and one incomplete predictor that covers only a subset of
all available genotypes.
Here, we explore the impact of which data are covered by the incomplete 
predictor on the predictive abilities obtained via single-step approaches by
focusing exclusively on the 149 inbred lines from the maize diversity panel
that are covered by both, genomic and transcriptomic information.
First, we assmbled nine different core sets, which we defined as subsets
of genotypes that maximize the average genetic distance between genotypes of
the full set.
The nine core sets differ in that each covers a different number of genotypes,
ranging from only 10\% to to 90\% of all genotypes covered by the incomplete
predictor (Fig. \ref{fig:CoreSetPCA}).

@@

Predictive abilities obtained with the different core sets varied widely within
traits (Fig. \ref{fig:InbredResults}C).
For the traits CW and PH, which were better predicted via G than via T in the 
reduced set of lines, predictive abilities were higher when only a small number 
of genotypes was included in the core set.
The latter corresponds to a situation where a large number of genotypes is
being imputed and the majority of the information is derived from genomic
information. 
For most other traits, an initial increase in predictive abilities when moving
from a core set size of 0.1 to 0.5/0.6 was observed, followed by a drop around
0.6/0.7 and a subsequent hike up to a core set size of 0.9.
A principal component analysis depicts the coverage of the genetic space by
genotypes (Fig. \ref{fig:PopStructure}).
With a core set size of 50-60\% of the 149 genotypes, the congruence between
core set and complement was high whereas genotypes in the south-west direction
of the PCA plot were overrepresented compared to genotypes in the east 
direction at a core set size of 0.7.

@@

Of the 149 genotypes, 17 belonged to more than 50\% to the first ancestral
population (A1), 109 belonged to more than 50\% to the second ancestral
population (A2) and 19 belonged to more than 50\% to the third ancestral
population while four genotypes could not be unequivocally assigned to either
ancestral population (mixed).
To adjust the size of the set of genotypes largely belonging to A2 to the other
two sets, we selected 19 genotypes from A2 using previously described core
sampling.
For single-step prediction, each of the three genotype sets were then declared
as having genomic and transcriptomic information whereas their complements were
covered by transcriptomic data only.
Genotypes from A1 and A3, which were at the far ends in the PCA plot (Fig.
\ref{fig:PopStructure}B), were estimated with less precision than genotypes
from the central cluster of genotypes belonging largely to the A2 population
(Fig. \ref{fig:PopStructure}A, Fig. \ref{fig:InbredResults}B).
